---
title: "Occurrence indicators: pre-processing"
author:
- Damiano Oldoni
- Toon Van Daele
- Tim Adriaens
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This document describes the pre-processing of occurrence data of alien species in Belgium. The outputs of this pipeline will be used as input for modelling.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, tidy='styler')
```

Load libraries:

```{r load_libraries, message = FALSE}
library(tidyverse) # To do data science
library(here) # To find files
library(lubridate) # To work with dates
library(rgbif) # To get taxa from publisehd unified checklist
library(elodea) # To work with GBIF checklists
```

# Get data

## Occurrence cubes

Read the occurrence cubes from `data/interim`.

First, the cube related to alien taxa:

```{r load_cube}
df <- readr::read_csv(
  here::here("data", "input", "be_alientaxa_cube.csv"),
  col_types = cols(
    year = col_double(),
    eeaCellCode = col_character(),
    taxonKey = col_double(),
    occurrences = col_double(),
    minCoordinateUncertaintyInMeters = col_double()
  ),
  na = ""
)
```

Second, the cube related to the baseline:

```{r load_baseline}
df_bl <- readr::read_csv(
  here::here("data", "input", "be_classes_cube.csv"),
  col_types = cols(
    year = col_double(),
    eeaCellCode = col_character(),
    classKey = col_double(),
    occurrences = col_double(),
    minCoordinateUncertaintyInMeters = col_double()
  ),
  na = ""
)
```

### Download

For users who want to reproduce the workflow from this stage, notice we are going to upload the cubes on the Zenodo deposit [10.5281/zenodo.3632749](https://doi.org/10.5281/zenodo.3632749): link to the the right version will follow here.

### Alien species

We remove columns we will never use:

```{r remove_coords_uncertainty}
df <-
  df %>%
  dplyr::select(-minCoordinateUncertaintyInMeters)
```

and rename column `occurrences` to `obs`:

```{r rename_n_to_obs}
df <-
  df %>%
  dplyr::rename(obs = occurrences)
```

### Baseline

We remove columns we will never use:

```{r remove_coords_uncertainty_baseline}
df_bl <-
  df_bl %>%
  dplyr::select(-minCoordinateUncertaintyInMeters)
```

and rename column `occurrences` to `cobs`:

```{r rename_n_to_obs_bl}
df_bl <-
  df_bl %>%
  dplyr::rename(cobs = occurrences)
```

## Checklist data

## Taxon core

We read input data based on [Global Register of Introduced and Invasive Species - Belgium](https://www.gbif.org/dataset/6d9e952f-948c-4483-9807-575348147c7e). We saved these taxa in previous pipeline in file `griis_taxon_core.tsv` in `data/input`:

```{r read_taxa}
taxa_df <- readr::read_tsv(
  here::here("data", "input", "griis_taxon_core.tsv"),
  na = ""
)
```

### Retrieve distribution information of alien species in Belgium

Retrieve the distribution information of the [Global Register of Introduced and Invasive Species - Belgium](https://www.gbif.org/dataset/6d9e952f-948c-4483-9807-575348147c7e) as it will be used to define some temporal restraints. We save this information in file `griis_distributions.tsv` in `data/output`:

```{r get_distributions, cache=TRUE}
distributions <- elodea::get_distributions(
  datasetKey ="6d9e952f-948c-4483-9807-575348147c7e"
)
```

## Filter data

### Cells in occurrence cubes not in EEA Belgian grid

The occurrence cubes were created based on GBIF occurrences filtered using country = `BE`. The cell codes were then created by manipulating the coordinates as strings, i.e. no spatial filter with the EEA grid was applied. So, it can happen that (few) cells in the cube are not linked with occurrences in Belgian territory as `country = BE` is in rare cases not enough. We remove them by using the geopackage containing the regional information of the EEA grid cells as shown in the following step. The regional information will be added at a later stage.

Read the EEA grid with regional information from `utm1_bel_with_regions.gpkg` in `data/output`:

```{r read_regions_info}
gpkg_path <- here::here("data", "output", "utm1_bel_with_regions.gpkg")
eea_grid_regions <- sf::st_read(gpkg_path)
```

Preview:

```{r preview_regions_info}
eea_grid_regions %>% head()
```

Notice that the column `isBelgium` is the union of the columns `isFlanders`, `isWallonia` and `isBrussels`. The 15km buffer zone and the maritime federal zones are not covered. At the moment we will not use the column `isBelgium` as occurrences on the maritime federal zones or in proximity of the coast would be otherwise excluded. Still, notice that these areas cover almost 40% of the EEA grid:

```{r count-n-cells-in-out-terestrial-BE}
eea_grid_regions %>%
  # Transform to data.frame
  as.data.frame() %>%
  # Count the number of cells in Belgium
  count(isBelgium) %>%
  # Add percentage as a column
  mutate(percentage = n / sum(n) * 100)
```

We identify the rows to remove in alien species occurrence cube, `df`, and  baseline, `df_bl`, by doing an `anti_join`:

```{r show-cells-out-of-belgium}
df %>%
  dplyr::filter(!eeaCellCode %in% eea_grid_regions$CELLCODE)
```

```{r show-cells-out-of-belgium-baseline}
df_bl %>%
  dplyr::filter(!eeaCellCode %in% eea_grid_regions$CELLCODE)
```

We remove them:

```{r remove-cells-out-of-be}
df <- df %>%
  dplyr::filter(eeaCellCode %in% eea_grid_regions$CELLCODE)
```

```{r remove-cells-out-of-be-baseline}
df_bl <- df_bl %>%
  dplyr::filter(!eeaCellCode %in% eea_grid_regions$CELLCODE)
```

### Taxa in unified checklist without occurrences

There is a group of taxa defined in the unified checklist which are not present in the occurrence data, i.e. the presence of these taxa, although confirmed by authoritative's checklists, is still not sustained by observations:

```{r taxa_without_occs}
alien_taxa_without_occs <-
  taxa_df %>%
  dplyr::filter(!acceptedKey %in% df$taxonKey) %>%
  dplyr::arrange(nubKey)
alien_taxa_without_occs
```

where `taxonKey` is the unique GBIF identifier of the taxon as published in the unified checklist, while `nubKey` is the unique GBIF identifier of the taxon in the GBIF Backbone Taxonomy. The field `acceptedTaxonKey` is equal to `nubKey` if `nubKey` is not a synonym. We save this data.frame as `alien_taxa_without_occurrences.tsv` in `data/output`:

```{r save_alien_taxa_without_occs}
write_tsv(
  alien_taxa_without_occs,
  here::here("data", "output", "alien_taxa_without_occs.tsv"),
  na = ""
)
```

## Regional information

Add regional information to the occurrence cubes by doing a left join with the EEA grid with regional information:

```{r add_regional_info}
df <- df %>%
  dplyr::inner_join(
    eea_grid_regions %>% as.data.frame(),
    by = c("eeaCellCode" = "CELLCODE")
  ) %>%
  dplyr::select(-c("EOFORIGIN", "NOFORIGIN", "geom"))
df_bl <- df_bl %>%
  dplyr::inner_join(
    eea_grid_regions %>% as.data.frame(),
    by = c("eeaCellCode" = "CELLCODE")
  ) %>%
  dplyr::select(-c("EOFORIGIN", "NOFORIGIN", "geom"))
```

## Grid cells intersecting protected areas

We get information about the intersection of grid cells with Belgian protected areas by reading file `intersect_EEA_ref_grid_protected_areas.tsv` in `data/output`:

```{r read_protected_areas}
df_prot_areas <- readr::read_tsv(
  here::here(
    "data",
    "interim",
    "intersect_EEA_ref_grid_protected_areas.tsv"
  ),
  na = ""
)
```

Preview:

```{r preview_prot_areas}
df_prot_areas %>%
  head()
```

We are interested in areas included in Natura2000. We remove columns we are not interested in:

```{r select_cols_prot_areas}
df_prot_areas <-
  df_prot_areas %>%
  dplyr::select(
    CELLCODE,
    natura2000
  )
```


#  Pre-processing

## Extract x-y coordinates from cellcodes

Extract coordinates from column `eeaCellCode`. For example, the coordinates of cellcode `1kmE3895N3116` are `x = 3895` and `y = 3116`:

```{r extract_x_y}
df_xy <-
  df %>%
  dplyr::distinct(eeaCellCode) %>%
  dplyr::bind_cols(
    dplyr::tibble(
      x = unlist(stringr::str_extract_all(unique(df$eeaCellCode),
        pattern = "(?<=E)[0-9\\-]+"
      )),
      y = unlist(stringr::str_extract_all(unique(df$eeaCellCode),
        pattern = "(?<=N)[0-9\\-]+"
      ))
    ) %>%
    dplyr::mutate(across(where(is.character), as.integer))
  )
```

Preview:

```{r x_y_preview}
df_xy %>% head()
```

Do the same for baseline data:

```{r extract_x_y_baseline}
df_bl_xy <-
  df_bl %>%
  dplyr::distinct(eeaCellCode) %>%
  bind_cols(
    tibble(
      x = unlist(str_extract_all(unique(df_bl$eeaCellCode),
        pattern = "(?<=E)[0-9\\-]+"
      )),
      y = unlist(str_extract_all(unique(df_bl$eeaCellCode),
        pattern = "(?<=N)[0-9\\-]+"
      ))
    ) %>%
      mutate_all(as.integer)
  )
```

Preview:

```{r x_y_preview_baseline}
df_bl_xy %>% head()
```

## Apply temporal and spatial restraints

From the distribution information, retrieve the first year of introduction (`first_observed`) for each taxon:

```{r get_first_observed_from_eventDate}
distributions <- distributions %>%
  # Parse eventDate to get year of introduction. Get the characters before `/` if present and convert to numberic if not character `"NA"`
  dplyr::mutate(
    first_observed = case_when(
      is.na(eventDate) ~ NA_real_,
      stringr::str_detect(eventDate, "/") ~ as.numeric(stringr::str_extract(eventDate, "^[0-9]+"))
    )
  )
```

Add such information to the taxonomic information:

```{r add_first_observed}
taxa_df <- taxa_df %>%
  dplyr::left_join(
    distributions %>% dplyr::select(taxonKey, first_observed),
    dplyr::join_by(taxonKey)
  )
```

Select alien species introduced after 1980 (`year_of_introduction`: 1950) or whose date of introduction is not known:

```{r select_aliens_after_year_of_introduction}
year_of_introduction <- 1950

recent_alien_species <-
  taxa_df %>%
  # Lower limit on date of first introduction if present
  dplyr::filter(first_observed >= year_of_introduction | is.na(first_observed))
```

Number of (accepted) taxa recently introduced or whose introduction year is not specified:

```{r n_recent_taxa}
length(unique(recent_alien_species$acceptedKey))
```

Alien species excluded as first introduction date is earlier than 1950:

```{r recent_alien_species}
old_introductions_taxa <-
  taxa_df %>%
  dplyr::filter(first_observed < year_of_introduction)
old_introductions_taxa
```

We save them in file `taxa_introduced_in_BE_before_1950.tsv` in `data/output`:

```{r save_old_introduced_taxa_as_tsv_file}
write_tsv(old_introductions_taxa,
  here::here(
    "data",
    "output",
    paste0(
      "taxa_introduced_in_BE_before_",
      year_of_introduction,
      ".tsv"
    )
  ),
  na = ""
)
```

Remove from the occurrence cubes the taxa introduced before 1950 and add x-y coordinates:

```{r apply_restraints_time_space_occ_data}
df <-
  df %>%
  # Recently introduced taxa
  dplyr::filter(taxonKey %in% recent_alien_species$acceptedKey) %>%
  dplyr::left_join(df_xy, by = "eeaCellCode")
```

Add x-y coordinates to the baseline cube:

```{r add_space_baseline_data}
df_bl_xy <-
  df_bl %>%
  dplyr::left_join(df_bl_xy, by = "eeaCellCode")
```

Notice that the occurrence cubes are already temporally limited to GBIF occurrences observed after 1980.

## Add information about presence in protected areas

We add whether the grid cell intersects any of the Natura2000 Belgian protected areas.

```{r add_natura2000_col}
df <-
  df %>%
  dplyr::left_join(df_prot_areas,
    by = c("eeaCellCode" = "CELLCODE")
  )
```

Preview:

```{r preview_df_with_natura2000_col}
df %>% head()
```

We do the same for baseline data:

```{r add_natura2000_col_baseline}
df_bl <-
  df_bl %>%
  dplyr::left_join(df_prot_areas,
    by = c("eeaCellCode" = "CELLCODE")
  )
```

Preview:

```{r preview_df_bl_with_natura2000_col}
df_bl %>% head()
```


# Save data

## Save enriched cubes

Save the cubes:

```{r save_data_as_csv}
readr::write_csv(
  df,
  file = here::here("data", "interim", "be_alientaxa_cube_enriched.csv"),
  na = ""
)
readr::write_csv(
  df_bl,
  file = here::here("data", "interim", "be_classes_cube_enriched.csv"), 
  na = ""
)
```

## Save distribution information of alien species in Belgium

Save distribution extension:

```{r save_distributions}
write_tsv(
  distributions,
  file = here::here("data", "output", "griis_distributions.tsv"),
  na = ""
)
```
