---
title: 'Detect appearing and reappearing species'
author:
- Damiano Oldoni
- Toon Van Daele
- Tim Adriaens
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    code_folding: show
---

In this document we analyze the Belgian datacube to detect possible (re)appearing taxa in the last year. Such detection is based on some basic rules instead of using GAM (Generalized Additive Models) as the number of GBIF observations decreases sensibly during the current year as effect of a publishing delay. We also take into account the appearing/reappearing in protected areas as additional warning.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, tidy='styler')
```

Load libraries:

```{r load_libraries, message = FALSE}
library(tidyverse) # To do data science
library(here) # To find files
library(rgbif) # To get information from GBIF
```

# Read data

## Read occurrence cubes

We read the enriched cubes:

```{r read_cubes, message = FALSE}
df <- readr::read_csv(
  here::here("data", "interim", "be_alientaxa_cube_enriched.csv"),
  na = ""
)
df_bl <- readr::read_csv(
  here::here("data", "interim", "be_classes_cube_enriched.csv"),
  na = ""
)
```

## Read taxonomic information

We read the taxonomic core from the GRIIS Belgium as done in previous pipelines:

```{r taxon_info}
taxa_df <- readr::read_tsv(
  here::here("data", "input", "griis_taxon_core.tsv"),
  na = ""
)
```

Preview:

```{r preview_taxa_df}
head(taxa_df)
```


# Detect appearing and reappearing species

For emerging trends in the next workflow we do not take into account the present year and the last year as data publication delays can occur. However, for getting a list of appearing and reappearing species we can of course consider the present year and the last year as evaluation years. 

```{r eval_years}
# First year to consider for evaluation of (re)appearing taxa
first_year <- lubridate::year(Sys.Date()) - 2 - 2

# First evaluation year
last_year <- lubridate::year(Sys.Date())

# Evaluation years
years <- seq(first_year, last_year)
```

## Appearing species

We define the function `get_appearing_taxa` for detecting appearing taxa (this function could be generalised and added to TrIAS package `trias` later):

```{r decision_rules_function, class.source = 'fold-hide'}
#' @param cube data frame with occurrence cube. Columns `taxonKey` and `taxonName` must be present.
#' @param first_eval_year first year to consider for evaluation of appearing taxa
#' @return data.frame with appearing taxa
get_appearing_taxa <- function(cube, first_eval_year) {
    cube %>%
      dplyr::group_by(taxonKey, taxonName) %>%
      dplyr::summarise(
        first_year = min(year),
        .groups = "drop"
      ) %>%
      dplyr::filter(first_year >= first_eval_year)
  }
```

### Appearing species in Belgium

Apply function defined above over the entire cube referring to Belgium:

```{r appearing_taxa_BE}
appearing_taxa_be <- get_appearing_taxa(
  cube = df,
  first_eval_year = first_year
  ) %>%
  dplyr::arrange(desc(first_year)) %>%
  dplyr::left_join(
    taxa_df %>%
      dplyr::select(acceptedKey, acceptedKingdom),
    dplyr::join_by(taxonKey == acceptedKey)
  )
```

Appearing taxa in Belgium:

```{r show_appearing_taxa_BE}
appearing_taxa_be
```

### Appearing species in Belgian protected areas

Apply `get_appearing_taxa()` only to the part of the cube referring to cells in protected areas:

```{r appearing_taxa_prot_areas}
appearing_taxa_prot_areas <- df %>%
  dplyr::filter(natura2000 == TRUE) %>%
  get_appearing_taxa(first_eval_year = first_year) %>%
  dplyr::arrange(desc(first_year)) %>%
  dplyr::left_join(
    taxa_df %>%
      dplyr::select(acceptedKey, acceptedKingdom),
    dplyr::join_by(taxonKey == acceptedKey)
  )
```

Appearing taxa in Belgian protected areas:

```{r show_appearing_taxa_prot_areas}
appearing_taxa_prot_areas
```

### Appearing species in Flanders

Apply `get_appearing_taxa()` only to the part of the cube referring to Flanders:

```{r appearing_taxa_flanders}
appearing_taxa_flanders <- df %>%
  dplyr::filter(isFlanders == TRUE) %>%
  get_appearing_taxa(first_eval_year = first_year) %>%
  dplyr::arrange(desc(first_year)) %>%
  dplyr::left_join(
    taxa_df %>%
      dplyr::select(acceptedKey, acceptedKingdom),
    dplyr::join_by(taxonKey == acceptedKey)
  )
```

Appearing taxa in Flanders:

```{r show_appearing_taxa_flanders}
appearing_taxa_flanders
```

### Appearing species in Flemish protected areas

Apply `get_appearing_taxa()` only to the part of the cube referring to cells in protected areas in Flanders:

```{r appearing_taxa_prot_areas_flanders}
appearing_taxa_prot_areas_flanders <- df %>%
  dplyr::filter(natura2000 == TRUE, isFlanders == TRUE) %>%
  get_appearing_taxa(first_eval_year = first_year) %>%
  dplyr::arrange(desc(first_year)) %>%
  dplyr::left_join(
    taxa_df %>%
      dplyr::select(acceptedKey, acceptedKingdom),
    dplyr::join_by(taxonKey == acceptedKey)
  )
```

Appearing taxa in Flemish protected areas:

```{r show_appearing_taxa_prot_areas_flanders}
appearing_taxa_prot_areas_flanders
```

### Appearing species in Wallonia

Apply `get_appearing_taxa()` only to the part of the cube referring to Wallonia:

```{r appearing_taxa_wallonia}
appearing_taxa_wallonia <- df %>%
  dplyr::filter(isWallonia == TRUE) %>%
  get_appearing_taxa(first_eval_year = first_year) %>%
  dplyr::arrange(desc(first_year)) %>%
  dplyr::left_join(
    taxa_df %>%
      dplyr::select(acceptedKey, acceptedKingdom),
    dplyr::join_by(taxonKey == acceptedKey)
  )
```

Appearing taxa in Wallonia:

```{r show_appearing_taxa_wallonia}
appearing_taxa_wallonia
```

### Appearing species in Walloon protected areas

Apply `get_appearing_taxa()` only to the part of the cube referring to cells in protected areas in Wallonia:

```{r appearing_taxa_prot_areas_wallonia}
appearing_taxa_prot_areas_wallonia <- df %>%
  dplyr::filter(natura2000 == TRUE, isWallonia == TRUE) %>%
  get_appearing_taxa(first_eval_year = first_year) %>%
  dplyr::arrange(desc(first_year)) %>%
  dplyr::left_join(
    taxa_df %>%
      dplyr::select(acceptedKey, acceptedKingdom),
    dplyr::join_by(taxonKey == acceptedKey)
  )
```

Appearing taxa in Walloon protected areas:

```{r show_appearing_taxa_prot_areas_wallonia}
appearing_taxa_prot_areas_wallonia
```

### Appearing species in Brussels

Apply `get_appearing_taxa()` only to the part of the cube referring to Brussels:

```{r appearing_taxa_brussels}
appearing_taxa_brussels <- df %>%
  dplyr::filter(isBrussels == TRUE) %>%
  get_appearing_taxa(first_eval_year = first_year) %>%
  dplyr::arrange(desc(first_year)) %>%
  dplyr::left_join(
    taxa_df %>%
      dplyr::select(acceptedKey, acceptedKingdom),
    dplyr::join_by(taxonKey == acceptedKey)
  )
```

Appearing taxa in Brussels:

```{r show_appearing_taxa_brussels}
appearing_taxa_brussels
```

### Appearing species in Brussels protected areas

Apply `get_appearing_taxa()` only to the part of the cube referring to cells in protected areas in Brussels:

```{r appearing_taxa_prot_areas_brussels}
appearing_taxa_prot_areas_brussels <- df %>%
  dplyr::filter(natura2000 == TRUE, isBrussels == TRUE) %>%
  get_appearing_taxa(first_eval_year = first_year) %>%
  dplyr::arrange(desc(first_year)) %>%
  dplyr::left_join(
    taxa_df %>%
      dplyr::select(acceptedKey, acceptedKingdom),
    dplyr::join_by(taxonKey == acceptedKey)
  )
```

Appearing taxa in Brussels protected areas:

```{r show_appearing_taxa_prot_areas_brussels}
appearing_taxa_prot_areas_brussels
```

## Reappearing species

We define the minimum latency (in years) before we speak about reappearance. For example, a latency time window of 5 years means that we consider a taxon reappearing in 2025 if observations occur in 2025 and 2019 or before:

```{r latency_years}
years_latency <- 5
```

We define the function `get_reappearing_taxa` for detecting reappearing taxa (this function could be generalised and added to TrIAS package `trias` later):

```{r get_reappearing_taxa_function, class.source = 'fold-hide'}
#' @param cube data frame with occurrence cube. Columns `taxonKey` and `taxonName` must be present.
#' @param first_eval_year first year to consider for evaluation of appearing taxa
#' @param latency_threshold minimum latency (in years) before we speak about reappearance. For example, a latency time window of 5 years means that we consider a taxon reappearing in 2025 if observations occur in 2025 and 2019 or before.
#' @return data.frame with reappearing taxa
get_reappearing_taxa <- function(
    cube,
    first_eval_year,
    latency_threshold) {
  reappearing_species <- cube %>%
    dplyr::group_by(taxonKey, taxonName) %>%
    dplyr::summarise(
      # set to min(year) from the first evaluation year if there are 
      # occurrences in those years, otherwise set to NA. This way we can 
      # filter for species that reappear in the evaluation years or later.
      reappearance_year = ifelse(
        any(year >= first_eval_year),
        min(year[year >= first_eval_year]),
        NA_integer_
      ),
      years_without_occurrences = ifelse(
        any(year >= first_eval_year) & any(year < first_eval_year),
        min(year[year >= first_eval_year]) - max(year[year < first_eval_year]),
        NA_integer_
      ),
      .groups = "drop"
    ) %>%
    dplyr::filter(
      !is.na(reappearance_year) & 
        !is.na(years_without_occurrences) & 
        years_without_occurrences >= latency_threshold
    )
}
```

### Reappearing species in Belgium

Apply function defined above over the entire cube referring to Belgium:

```{r reappearing_taxa_BE}
reappearing_taxa_be <- get_reappearing_taxa(
  cube = df,
  first_eval_year = first_year,
  latency_threshold = years_latency
  ) %>%
  dplyr::arrange(desc(reappearance_year), desc(years_without_occurrences)) %>%
  dplyr::left_join(
    taxa_df %>%
      dplyr::select(acceptedKey, acceptedKingdom),
    dplyr::join_by(taxonKey == acceptedKey)
  )
```

Reappearing taxa in Belgium:

```{r show_reappearing_taxa_BE}
reappearing_taxa_be
```

### Reappearing species in Belgian protected areas

Apply `get_reappearing_taxa()` only to the part of the cube referring to cells in protected areas:

```{r reappearing_taxa_prot_areas}
reappearing_taxa_prot_areas <- df %>%
  dplyr::filter(natura2000 == TRUE) %>%
  get_reappearing_taxa(
    first_eval_year = first_year,
    latency_threshold = years_latency
  ) %>%
  dplyr::arrange(desc(reappearance_year), desc(years_without_occurrences)) %>%
  dplyr::left_join(
    taxa_df %>%
      dplyr::select(acceptedKey, acceptedKingdom),
    dplyr::join_by(taxonKey == acceptedKey)
  )
```

Reappearing taxa in Belgian protected areas:

```{r show_reappearing_taxa_prot_areas}
reappearing_taxa_prot_areas
```

### Reappearing species in Flanders

Apply `get_reappearing_taxa()` only to the part of the cube referring to Flanders:

```{r reappearing_taxa_flanders}
reappearing_taxa_flanders <- df %>%
  dplyr::filter(isFlanders == TRUE) %>%
  get_reappearing_taxa(
    first_eval_year = first_year,
    latency_threshold = years_latency
  ) %>%
  dplyr::arrange(desc(reappearance_year), desc(years_without_occurrences)) %>%
  dplyr::left_join(
    taxa_df %>%
      dplyr::select(acceptedKey, acceptedKingdom),
    dplyr::join_by(taxonKey == acceptedKey)
  )
```

Reappearing taxa in Flanders:

```{r show_reappearing_taxa_flanders}
reappearing_taxa_flanders
```

### Reappearing species in Flemish protected areas

Apply `get_reappearing_taxa()` only to the part of the cube referring to cells in protected areas in Flanders:

```{r reappearing_taxa_prot_areas_flanders}
reappearing_taxa_prot_areas_flanders <- df %>%
  dplyr::filter(natura2000 == TRUE, isFlanders == TRUE) %>%
  get_reappearing_taxa(
    first_eval_year = first_year,
    latency_threshold = years_latency
  ) %>%
  dplyr::arrange(desc(reappearance_year), desc(years_without_occurrences)) %>%
  dplyr::left_join(
    taxa_df %>%
      dplyr::select(acceptedKey, acceptedKingdom),
    dplyr::join_by(taxonKey == acceptedKey)
  )
```

Reappearing taxa in Flemish protected areas:

```{r show_reappearing_taxa_prot_areas_flanders}
reappearing_taxa_prot_areas_flanders
```

### Reappearing species in Wallonia

Apply `get_appearing_taxa()` only to the part of the cube referring to Wallonia:

```{r reappearing_taxa_wallonia}
reappearing_taxa_wallonia <- df %>%
  dplyr::filter(isWallonia == TRUE) %>%
  get_reappearing_taxa(
    first_eval_year = first_year,
    latency_threshold = years_latency
  ) %>%
  dplyr::arrange(desc(reappearance_year), desc(years_without_occurrences)) %>%
  dplyr::left_join(
    taxa_df %>%
      dplyr::select(acceptedKey, acceptedKingdom),
    dplyr::join_by(taxonKey == acceptedKey)
  )
```

Reappearing taxa in Wallonia:

```{r show_reappearing_taxa_wallonia}
reappearing_taxa_wallonia
```

### Reappearing species in Walloon protected areas

Apply `get_reappearing_taxa()` only to the part of the cube referring to cells in protected areas in Wallonia:

```{r reappearing_taxa_prot_areas_wallonia}
reappearing_taxa_prot_areas_wallonia <- df %>%
  dplyr::filter(natura2000 == TRUE, isWallonia == TRUE) %>%
  get_reappearing_taxa(
    first_eval_year = first_year,
    latency_threshold = years_latency
  ) %>%
  dplyr::arrange(desc(reappearance_year), desc(years_without_occurrences)) %>%
  dplyr::left_join(
    taxa_df %>%
      dplyr::select(acceptedKey, acceptedKingdom),
    dplyr::join_by(taxonKey == acceptedKey)
  )
```

Reappearing taxa in Walloon protected areas:

```{r show_reappearing_taxa_prot_areas_wallonia}
reappearing_taxa_prot_areas_wallonia
```

### Reappearing species in Brussels

Apply `get_appearing_taxa()` only to the part of the cube referring to Brussels:

```{r reappearing_taxa_brussels}
reappearing_taxa_brussels <- df %>%
  dplyr::filter(isBrussels == TRUE) %>%
  get_reappearing_taxa(
    first_eval_year = first_year,
    latency_threshold = years_latency
  ) %>%
  dplyr::arrange(desc(reappearance_year), desc(years_without_occurrences)) %>%
  dplyr::left_join(
    taxa_df %>%
      dplyr::select(acceptedKey, acceptedKingdom),
    dplyr::join_by(taxonKey == acceptedKey)
  )
```

Reappearing taxa in Brussels:

```{r show_reappearing_taxa_brussels}
reappearing_taxa_brussels
```

### Reappearing species in Brussels protected areas

Apply `get_reappearing_taxa()` only to the part of the cube referring to cells in protected areas in Brussels:

```{r reappearing_taxa_prot_areas_brussels}
reappearing_taxa_prot_areas_brussels <- df %>%
  dplyr::filter(natura2000 == TRUE, isBrussels == TRUE) %>%
  get_reappearing_taxa(
    first_eval_year = first_year,
    latency_threshold = years_latency
  ) %>%
  dplyr::arrange(desc(reappearance_year), desc(years_without_occurrences)) %>%
  dplyr::left_join(
    taxa_df %>%
      dplyr::select(acceptedKey, acceptedKingdom),
    dplyr::join_by(taxonKey == acceptedKey)
  )
```

Reappearing taxa in Brussels protected areas:

```{r show_reappearing_taxa_prot_areas_brussels}
reappearing_taxa_prot_areas_brussels
```

# Save data

## Appearing taxa

Save appearing taxa as tab separated files:

```{r save_appearing_taxa}
appearing_taxa <- list(
  appearing_taxa_be = appearing_taxa_be,
  appearing_taxa_prot_areas = appearing_taxa_prot_areas,
  appearing_taxa_flanders = appearing_taxa_flanders,
  appearing_taxa_prot_areas_flanders = appearing_taxa_prot_areas_flanders,
  appearing_taxa_wallonia = appearing_taxa_wallonia,
  appearing_taxa_prot_areas_wallonia = appearing_taxa_prot_areas_wallonia,
  appearing_taxa_brussels = appearing_taxa_brussels,
  appearing_taxa_prot_areas_brussels = appearing_taxa_prot_areas_brussels
)
purrr::iwalk(
  appearing_taxa,
  ~ write_tsv(
    .x,
    file = here::here("data", "output", paste0(.y, ".tsv")),
    na = ""
  )
)
```

## Reappearing taxa

Save reappearing taxa as tab separated files:

```{r save_reappearing_taxa}
reappearing_taxa_list <- list(
  reappearing_taxa_be = reappearing_taxa_be,
  reappearing_taxa_prot_areas = reappearing_taxa_prot_areas,
  reappearing_taxa_flanders = reappearing_taxa_flanders,
  reappearing_taxa_prot_areas_flanders = reappearing_taxa_prot_areas_flanders,
  reappearing_taxa_wallonia = reappearing_taxa_wallonia,
  reappearing_taxa_prot_areas_wallonia = reappearing_taxa_prot_areas_wallonia,
  reappearing_taxa_brussels = reappearing_taxa_brussels,
  reappearing_taxa_prot_areas_brussels = reappearing_taxa_prot_areas_brussels
)
purrr::iwalk(
  reappearing_taxa_list,
  ~ write_tsv(
    .x,
    file = here::here("data", "output", paste0(.y, ".tsv")),
    na = ""
  )
)
```
