---
title: 'Emerging status: modelling'
author:
- Damiano Oldoni
- Toon Van Daele
- Soria Delva
- Tim Adriaens
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    code_folding: hide
---

This document describes the modelling to assess the emerging status of alien species.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, result = FALSE, tidy='styler')
```

Load libraries:

```{r load_libraries, message = FALSE}
library(tidyverse) # To do data science
library(tidyselect) # To help tidyverse functions
library(dplyr) # To provide feedback on dplyr functions
library(here) # To find files
library(trias)
library(INBOtheme) # To load graphic INBO theme
```

# Read data

## Read occurrence cubes

We read the cubes with taxonomic and spatial information (regions and protected areas):

```{r read_cubes, message = FALSE}
df <- readr::read_csv(
  here::here("data", "interim", "be_alientaxa_cube_enriched.csv"),
  na = ""
)
df_bl <- readr::read_csv(
  here::here("data", "interim", "be_classes_cube_enriched.csv"),
  na = ""
)
```

Preview the cube: 

```{r}
head(df)
```

Preview the baseline cube:

```{r}
head(df_bl)
```

## Read taxonomic information

We read the taxonomic core from the GRIIS Belgium as done in previous pipelines:

```{r taxon_info}
taxa_df <- readr::read_tsv(
  here::here("data", "input", "griis_taxon_core.tsv"),
  na = ""
)
```

Preview:

```{r preview_taxa_df}
head(taxa_df)
```


We set `isBelgium` = `TRUE` for all grid cells. Notice that at this moment `isBelgium` = `FALSE` for North Sea areas:

```{r add_isBelgium}
df$isBelgium <- TRUE
df_bl$isBelgium <- TRUE
```

Add class information to the cube:

```{r add_class}
df <- df %>%
  dplyr::left_join(
    taxa_df %>% dplyr::distinct(acceptedKey, acceptedClassKey, acceptedClass, acceptedKingdom),
     dplyr::join_by(taxonKey == acceptedKey)
  ) %>%
  dplyr::rename(
    classKey = acceptedClassKey,
    class = acceptedClass,
    kingdom = acceptedKingdom
  )
```

Preview:

```{r}
head(df)
```

## Add information about regional protected areas

We would like to add information about regional protected areas, i.e. grid cells intersecting protected areas in Flanders, Wallonia and Brussels. We use the columns `natura2000`, `isFlanders`, `isWallonia` and `isBrussels`:

```{r add_region_protectedareas}
df <- df %>%
  dplyr::mutate(
    isFlandersNatura2000 = isFlanders == TRUE & natura2000 == TRUE,
    isWalloniaNatura2000 = isWallonia == TRUE & natura2000 == TRUE,
    isBrusselsNatura2000 = isBrussels == TRUE & natura2000 == TRUE
  ) %>%
  dplyr::relocate(dplyr::ends_with("natura2000"), .before = x)
```

Preview:

```{r}
head(df)
```

Same for cube at class level (baseline):

```{r add_region_protectedareas_class}
df_bl <- df_bl %>%
  dplyr::mutate(
    isFlandersNatura2000 = isFlanders == TRUE & natura2000 == TRUE,
    isWalloniaNatura2000 = isWallonia == TRUE & natura2000 == TRUE,
    isBrusselsNatura2000 = isBrussels == TRUE & natura2000 == TRUE
  ) %>%
  dplyr::relocate(dplyr::ends_with("natura2000"), .before = x)
```

Preview:

```{r}
df_bl
```

# Create time series for number of occurrences and number of occupied grid cells

Define function to calculate number of occurrences and grid cells for a specific area (Belgium, protected areas or Belgian regions):

```{r lump_geographic_info}
#' Function to calculate the number of observations and occupied cells at taxon 
#' level per year for a specific area.
#' 
#' @param cube Data frame with enriched cube at taxon-level.
#' @param cube_bl Data frame with enriched cube at class-level.
#' @param area Character. Name of the column in the cube to filter by (e.g.
#' "isBelgium", "isFlanders", "natura2000"). Default: `NULL`: no filtering by
#' area.
calculate_variables <- function(cube, cube_bl, area = NULL) {
  # Filter by area
  if (!is.null(area)) {
    cube <- cube %>%
    dplyr::filter(!!sym(area) == TRUE)
  }
  
  cube_bl <-  cube_bl %>%
    dplyr::filter(!!sym(area) == TRUE)
  # Calculate the number of observations and occupied cells at taxon level per
  # year
  variables <- cube %>%
    dplyr::group_by(taxonKey, taxonName, year, class, classKey, kingdom) %>%
    dplyr::summarise(
      obs = sum(obs),
      ncells = dplyr::n(),
      .groups = "drop"
    )
  # Calculate the number of observations and occupied cells at class level per
  # year
  variables_bl <- cube_bl %>%
    dplyr::group_by(year, classKey, class) %>%
    dplyr::summarise(
      cobs = sum(cobs),
      c_ncells = dplyr::n(),
      .groups = "drop"
    )
  # Add number of observations and occupied cells at class level to the
  # taxon-level data
  variables <- variables %>%
    dplyr::left_join(variables_bl, dplyr::join_by(year, classKey, class))
  # Reduce it by number of obs/cells of the alien taxon. If the number of
  # obs/cells of the alien taxon is higher than the number of obs/cells at class
  # level, we set the value to 0 (we cannot have negative values)
  variables <- variables %>%
    dplyr::mutate(
      cobs = cobs - obs,
      c_ncells = c_ncells - ncells
    ) %>%
    dplyr::mutate(
      cobs = ifelse(cobs < 0, 0, cobs),
      c_ncells = ifelse(c_ncells < 0, 0, c_ncells)
    )
  # Add 0s for years with no observations or occupied cells for a taxon
  variables <- variables %>%
    group_by(taxonKey, taxonName, class, classKey, kingdom) %>%
    tidyr::complete(year = tidyr::full_seq(year, 1), fill = list(obs = 0, ncells = 0, cobs = 0, c_ncells = 0)) %>%
    ungroup()
  return(variables)
}
```

Apply the function to the regions and protected areas:
```{r calc_variables_per_area}
areas <- c("isBelgium",
           "isFlanders",
           "isWallonia",
           "isBrussels",
           "natura2000",
           "isFlandersNatura2000",
           "isWalloniaNatura2000",
           "isBrusselsNatura2000"
)
names(areas) <- areas
df_ts <- purrr::map(
  areas,
  function(x) calculate_variables(df, df_bl, area = x)
)
```

We have now compact time series of the variables for all regions and the protected areas:

```{r preview_ts}
names(df_ts)
```

Previews:

```{r previews_compact_ts}
purrr::map(df_ts, head)
```

# Modelling

In this section we evaluate the emerging status by applying a decision rule strategy or, where possible, a statistical model called GAM (Generalized Additive Models).
For each evaluation year (see below) the output of both models is one of the following emerging status codes:

- `3`: emerging
- `2`: potentially emerging
- `1`: unclear
- `0`: not emerging

See documentation of [`trias::apply_decision_rules`](https://trias-project.github.io/trias/reference/apply_decision_rules.html) and [`trias::apply_gam()`](https://trias-project.github.io/trias/reference/apply_gam.html) for more information.

## Define evaluation period 

We define the time window (in years) we want to assess the emerging status:

```{r eval_years}
# Last evaluation year
last_year <- lubridate::year(Sys.Date()) - 2

# First evaluation year
first_year <- last_year - 2

# Evaluation years
evaluation_years <- seq(first_year, last_year)
evaluation_years
```

We remove recent data due to publishing delay. Underestimation of the number of observations would sensibly affect the GAM output.

```{r remove_recent_data}
df_ts <- purrr::map(
  df_ts, 
  function(df) df %>% dplyr::filter(year <= last_year)
)
```

## Remove taxa with time series missing the evaluation years

Some taxa have time series ending before the evaluation years, i.e. no data are in `last_year`. For these taxa, no trend can be assessed in the evaluation period, and thus they should be removed from the analysis. We identify them by checking if the last year with observations or occupied cells is before the first evaluation year:

```{r taxa_to_remove_no_data_in_evaluation_year}
taxa_to_remove <- purrr::map(
  df_ts,
  function(df) {
    df %>%
      dplyr::group_by(taxonKey, taxonName) %>%
      dplyr::summarise(
        last_year_obs = max(year[obs > 0], na.rm = TRUE),
        last_year_ncells = max(year[ncells > 0], na.rm = TRUE)
      ) %>%
      dplyr::filter(last_year_obs < last_year & last_year_ncells < last_year)
  }
)
taxa_to_remove
```

Remove them:

```{r remove_taxa_to_not_evaluate}
df_ts <- purrr::map2(
  df_ts, taxa_to_remove,
  function(df, t) {
    df %>%
      dplyr::filter(!taxonKey %in% t$taxonKey)
  }
)
```


## Remove appearing taxa and reappearing taxa

Taxa appearing or reappearing from the evaluation years up to now should be removed as no trend can be assesed for them. See [analysis of appearing and reappearing taxa](https://trias-project.github.io/indicators/06_occurrence_indicators_appearing_taxa.html). 

### Remove appearing taxa

We read the appearing taxa from the generated files in `data/output`:

```{r appearing_taxa}
dir_path <- here::here("data", "output")
appearing_taxa_files <- list(
  isBelgium = here::here(dir_path, "appearing_taxa_be.tsv"),
  isFlanders = here::here(dir_path, "appearing_taxa_flanders.tsv"),
  isWallonia = here::here(dir_path, "appearing_taxa_wallonia.tsv"),
  isBrussels = here::here(dir_path, "appearing_taxa_brussels.tsv"),
  natura2000 = here::here(dir_path, "appearing_taxa_prot_areas.tsv"),
  isFlandersNatura2000 = here::here(
    dir_path, "appearing_taxa_prot_areas_flanders.tsv"
    ),
  isWalloniaNatura2000 = here::here(
    dir_path, "appearing_taxa_prot_areas_wallonia.tsv"
    ),
  isBrusselsNatura2000 = here::here(
    dir_path, "appearing_taxa_prot_areas_brussels.tsv"
    )
)
appearing_taxa <- purrr::map(
  appearing_taxa_files,
  readr::read_tsv,
  na = ""
)
```

We remove them from the time series:

```{r remove_appearing_taxa_from_ts}
df_ts <- purrr::map2(
  df_ts, appearing_taxa,
  function(ts, taxa_to_remove) {
    ts %>%
      dplyr::filter(!taxonKey %in% taxa_to_remove$taxonKey)
  }
)
```

### Reappearing taxa

Similar to appearing taxa, we read the reappearing taxa from the generated files in `data/output`:

```{r reappearing_taxa}
dir_path <- here::here("data", "output")
reappearing_taxa_files <- list(
  isBelgium = here::here(dir_path, "reappearing_taxa_be.tsv"),
  isFlanders = here::here(dir_path, "reappearing_taxa_flanders.tsv"),
  isWallonia = here::here(dir_path, "reappearing_taxa_wallonia.tsv"),
  isBrussels = here::here(dir_path, "reappearing_taxa_brussels.tsv"),
  natura2000 = here::here(dir_path, "reappearing_taxa_prot_areas.tsv"),
  isFlandersNatura2000 = here::here(
    dir_path, "reappearing_taxa_prot_areas_flanders.tsv"
    ),
  isWalloniaNatura2000 = here::here(
    dir_path, "reappearing_taxa_prot_areas_wallonia.tsv"
    ),
  isBrusselsNatura2000 = here::here(
    dir_path, "reappearing_taxa_prot_areas_brussels.tsv"
    )
)
reappearing_taxa <- purrr::map(
  reappearing_taxa_files,
  readr::read_tsv,
  na = ""
)
```

We remove them from the time series:

```{r remove_reappearing_taxa_from_ts}
df_ts <- purrr::map2(
  df_ts, reappearing_taxa,
  function(ts, taxa_to_remove) {
    ts %>%
      dplyr::filter(!taxonKey %in% taxa_to_remove$taxonKey)
  }
)
```

## Decision rules

We define and apply some decision rules for occupancy and observations in all Belgium and in protected areas using function `apply_decision_rules()` from project package `trias`.

### Observations

We apply function `apply_decision_rules()` to the observations:

```{r apply_decision_rules_occs, message=FALSE}
# Define wrap function to apply decision rules for each evaluation year
apply_decision_rules_for_ev_years <- function(df, y_var, eval_years, taxa_col) {
  purrr::map_dfr(
    eval_years,
    function(year) {
      trias::apply_decision_rules(
        df = df,
        y_var = y_var,
        eval_year = year,
        taxonKey = taxa_col,
      )
    }
  )
}

# Apply the wrap function to BE, regions and protected areas for observations
em_decision_rules_occs <- purrr::map(
  df_ts,
  apply_decision_rules_for_ev_years,
  y_var = "obs",
  eval_years = evaluation_years,
  taxa_col = "taxonKey"
)
```

Previews:

```{r example_dec_rules_occs}
purrr::map(
  areas,
  function(x) {
    em_decision_rules_occs %>% 
      purrr::pluck(x) %>% 
      dplyr::slice_head(n = 10)
  }
)
```

### Occupancy

We apply function `trias::apply_decision_rules()` to occupancy:

```{r apply_decision_rules_occupancy, message=FALSE}
# Apply the wrap function to BE, regions and protected areas for occupancy
em_decision_rules_occupancy <- purrr::map(
  df_ts,
  apply_decision_rules_for_ev_years,
  y_var = "ncells",
  eval_years = evaluation_years,
  taxa_col = "taxonKey"
)
```

Previews:

```{r example_dec_rules_occupancy}
purrr::map(
  areas,
  function(x) {
    em_decision_rules_occupancy %>% 
      purrr::pluck(x) %>% 
      dplyr::slice_head(n = 10)
  }
)
```

## Generalized additive model (GAM)

We apply GAM to observations and occupancy in all Belgium, its regions and the protected areas using function `trias::apply_gam()` from project package `trias`.

Plots are saved in `./data/output/GAM_outputs` directory:

```{r plots_dir}
dir_name_basic <- here::here("data", "output", "GAM_outputs")
```

We also define the plot dimensions in pixels:

```{r plots_dimension}
plot_dimensions <- list(width = 2800, height = 1500)
```

We define also a wrap-up function, `apply_gam_for_eval_years` to apply GAM for each evaluation year and taxon:

```{r apply_gam_for_eval_years}
# Define wrap function to apply GAM for every time series (Belgium, its regions
# and protected areas)
apply_gam_for_eval_years <- function(df,
                                     region,
                                     eval_years,
                                     y_var,
                                     indicator,
                                     baseline_var,
                                     root_dir,
                                     y_label,
                                     w,
                                     h) {
  # Remove "is" from "isBrussels", "isFlanders", "isWallonia", etc. to create the
  # subdirectory to be added to the basic directory
  region <- stringr::str_remove(region, "^is")
  dir_name <- here::here(dir_name_basic, region)
  # Check if directory exists before creating
  if (!dir.exists(dir_name)) {
    dir.create(dir_name)
  }
  taxon_keys_names <- df %>%
    dplyr::distinct(taxonKey, taxonName)
  message(paste0("Applying GAM. Variable: ", y_var, "; region: ", region, "."))
  gam_occs <- map2(
    taxon_keys_names$taxonKey, taxon_keys_names$taxonName,
    function(t, n) {
      df_key <- df %>%
        dplyr::filter(taxonKey == t)
      class_key <- unique(df_key[["classKey"]])
      # Use covariate only if taxon belongs to a class (classKey is available)
      if (!is.na(class_key)) {
        results_gam <- trias::apply_gam(
          df = df_key,
          y_var = y_var,
          taxonKey = "taxonKey",
          eval_years = eval_years,
          type_indicator = indicator,
          baseline_var = baseline_var,
          taxon_key = t,
          name = n,
          df_title = region,
          dir_name = dir_name,
          y_label = y_label,
          saveplot = TRUE,
          width = w,
          height = h
        )
      } else {
        # If taxon doesn't belong to any class (classKey is not available), use
        # the default GAM model without covariate
        results_gam <- trias::apply_gam(
          df = df_key,
          y_var = y_var,
          taxonKey = "taxonKey",
          eval_years = eval_years,
          type_indicator = indicator,
          taxon_key = t,
          name = n,
          df_title = region,
          dir_name = dir_name,
          y_label = y_label,
          saveplot = TRUE,
          width = w,
          height = h
        )
      }
      return(results_gam)
    }
  )
  names(gam_occs) <- taxon_keys_names$taxonKey
  gam_occs
}
```

### Observations

We apply GAM modelling by using the wrap-up function defined above to observations in Belgium, its regions and protected areas. This step can take long.

```{r apply_gam_to_all_taxa, message=FALSE, cache=TRUE}
# Apply `apply_gam_for_eval_years` to observations:
gam_occs <- purrr::imap(
  df_ts,
  apply_gam_for_eval_years,
  eval_years = evaluation_years,
  y_var = "obs",
  indicator = "observations",
  baseline_var = "cobs",
  root_dir = dir_name_basic,
  y_label = "number of observations",
  w = plot_dimensions$width,
  h = plot_dimensions$height
)
```

Show results for _Pinus sylvestris_:

```{r show_results_preview_occs_all_regions}
taxon_example <- "5285637"
purrr::map(gam_occs, function(x) {
  x %>%
    purrr::pluck(taxon_example) %>%
    purrr::pluck("plot")
})
```


### Occupancy

We apply GAM modelling by using the wrap-up function defined above to occupancy in Belgium, its regions and protected areas. This step can take long.

```{r apply_gam_occupancy_to_all_taxa, message=FALSE, cache=TRUE}
gam_occupancy <- purrr::imap(
  df_ts,
  apply_gam_for_eval_years,
  eval_years = evaluation_years,
  y_var = "ncells",
  indicator = "occupancy",
  baseline_var = "c_ncells",
  root_dir = dir_name_basic,
  y_label = "measured occupancy (km2)",
  w = plot_dimensions$width,
  h = plot_dimensions$height
)
```

Show results for _Pinus Sylvestris_:

```{r preview_gam_occupancy_Belgium}
purrr::map(gam_occupancy, function(x) {
  x %>%
    purrr::pluck(taxon_example) %>%
    purrr::pluck("plot")
})
```


# Save results

## Decision rules

Save emerging status based on decision rules:

```{r save_decision_rules_results}
# Define function to save the results for Belgium, its regions and protected
# areas
save_decision_rules_output <- function(df, region, indicator) {
  # Remove "is" from "isBrussels", "isFlanders", "isWallonia" for a better filename
  region <- stringr::str_remove(region, "^is")
  readr::write_tsv(x = df,
    file = here::here(
      "data", "output",
      "decision_rules_outputs",
      paste0("output_decision_rules_", indicator, "_", region, ".tsv")
    ),
    na = ""
  )
}

# Apply function to save decision rules outputs applied to occurrences
purrr::iwalk(em_decision_rules_occs, function(df, region) {
  save_decision_rules_output(df, region, indicator = "occs")
})

# Apply function to save decision rules outputs applied to occupancy
purrr::iwalk(em_decision_rules_occupancy, function(df, region) {
  save_decision_rules_output(df, region, indicator = "occupancy")
})
```


## GAM models

Save complete outputs and summaries:

```{r  save_gam_outputs}
# Define function to save GAM model outputs and summaries for Belgium, its
# regions and protected areas
save_gam_output <- function(gam_results,
                            region,
                            indicator) {
  # Remove "is" from "isBrussels", "isFlanders", "isWallonia" for a better filename
  region <- stringr::str_remove(region, "^is")
  # Save outputs
  readr::write_tsv(
    # Merge the outputs of all taxa as a data.frame for saving
    purrr::map_dfr(gam_results, 
                   function(x) {
                     x$output
                    }),
    na = "",
    file = here::here(
      dir_name_basic,
      paste0("output_GAM_", indicator, "_", region, ".tsv")
    )
  )
  
  # Save summaries
  readr::write_tsv(
    # Merge the summaries of all taxa as a data.frame for saving
    purrr::map_dfr(gam_results, 
                   function(x) {
                     x$em_summary
                    }),
    na = "",
    file = here::here(
      dir_name_basic,
      paste0("summary_GAM_", indicator, "_", region, ".tsv")
    )
  )
}

# Apply function to save GAM outputs applied to ocurrences
purrr::iwalk(gam_occs, function(df, region) {
  save_gam_output(df, region, indicator = "occs")
})

# Apply function to save decision rules outputs applied to occupancy
purrr::iwalk(gam_occupancy, function(df, region) {
  save_gam_output(df, region, indicator = "occupancy")
})
```

## Plots

We create a zip file with all generated GAM plots:

```{r create-zip-file, results='hide'}
folder_path <- here::here("data", "output", "GAM_outputs")

# Get relative paths of PNG files
png_files <- list.files(folder_path,
                        pattern = "\\.png$",
                        recursive = TRUE)

# Change working directory temporarily
old_wd <- getwd()
setwd(folder_path)

# Create zip file
zip(zipfile = "GAM_plots.zip", files = png_files)

# Restore working directory
setwd(old_wd)

# Move zip file to the desired location (if needed)
file.rename(file.path(folder_path, "GAM_plots.zip"),
            here::here("data", "output", "GAM_outputs", "GAM_plots.zip"))
```

