---
title: 'Ranking species by emerging status'
author:
- Damiano Oldoni
- Toon Van Daele
- Tim Adriaens
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    code_folding: hide
---

This document follows the [modelling pipeline](https://trias-project.github.io/indicators/07_occurrence_indicators_modelling.html) to produce a list of introduced species ranked by emerging status.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      tidy='styler'
)
```

Load libraries:

```{r load_libraries}
library(tidyverse) # To do data science
library(tidylog) # To provide feedback on dplyr functions
library(here) # To find files
```

# Get data

## GAM outputs

We read the emerging status based on GAM, saved as one of the outputs of [modelling pipeline](https://trias-project.github.io/indicators/07_occurrence_indicators_modelling.html). We do it for Belgium, its regions (Flanders, Wallonia and Brussels) and protected areas and for both occurrences and observed occupancy (number of occupied cells):

```{r read_em_summary}
regions <- list("Belgium", 
                "Flanders", 
                "Wallonia", 
                "Brussels", 
                "natura2000_Belgium",
                "natura2000_Wallonia",
                "natura2000_Flanders",
                "natura2000_Brussels"
)
names(regions) <- regions
indicators <- c("occs", "occupancy")
names(indicators) <- indicators

# Create all combinations regions/indicators
combinations <- expand_grid(region = regions, indicator = indicators)

# Apply a function to read files for each combination
em_gam_list <- pmap(combinations, function(region, indicator) {
  paste("Reading", indicator, "for", region)
  readr::read_tsv(
        file = here::here(
          "data",
          "output",
          "GAM_outputs",
          paste0("summary_GAM_", indicator, "_", region, ".tsv" )),
        na = "") %>%
      tidylog::mutate(indicator =  paste(indicator, region, sep= "_"),
                      model = "GAM")
})

# Name the results
names(em_gam_list) <- paste("gam", combinations$indicator, 
                            combinations$region, 
                            sep = "_"
)
```

Preview GAM ouput for modelling the number of occurrences in Belgium (100 rows):

```{r preview_gam_occs_BE}
em_gam_list$gam_occs_Belgium %>% head(100)
```

Preview GAM ouput for modelling occupancy (number of 1x1km cells) in Belgium (100 rows):

```{r preview_gam_occupancy_BE}
em_gam_list$gam_occupancy_Belgium %>% head(100)
```

The emerging status (column `em_status`) score is encoded as follows:

- 3: emerging
- 2: potentially emerging
- 1: unclear
- 0: not emerging

As you can see in preview above, GAM could not be applied to all taxa (`em_status` = `NA`). Number of taxon/year combinations without emerging status:

```{r get_number_nas}
em_gam_na <- purrr::map(
  em_gam_list,
  function(df) {
    nas <- df %>%
      dplyr::filter(is.na(em_status)) %>%
      nrow()
    total <- nrow(df)
    list("NAs" = nas, "Total" = total)
  }
)
purrr::iwalk(
  em_gam_na, 
  function(x, y) {
    message(paste0("Number of NAs (", y, "): ", x$NAs, "/", x$Total))
  }
)
```

In these cases we will use the emerging status as assessed by applying decision rules.

## Decision rules outputs

We read the emerging status based on decision rules, saved as one of the outputs of [modelling pipeline](https://trias-project.github.io/indicators/07_occurrence_indicators_modelling.html):

```{r read_decision_rules_results}
# Apply a function to read files for each combination
em_decision_rules_list <- pmap(combinations, function(region, indicator) {
  paste("Reading", indicator, "for", region)
  readr::read_tsv(
        file = here::here(
          "data",
          "output",
          "decision_rules_outputs",
          paste0("output_decision_rules_", indicator, "_", region, ".tsv" )),
        na = "") %>%
      tidylog::mutate(indicator = paste(indicator, region, sep = "_"),
                      model = "decision_rules")
})

# Name the results
names(em_decision_rules_list) <- paste(
  "decision_rules",
  combinations$indicator,
  combinations$region,
  sep = "_"
)
```

Previewing output of decision rules for occupancy in Natura2000 protected areas (100 rows):

```{r preview_dr_list}
em_decision_rules_list$decision_rules_occupancy_natura2000_Belgium %>% head(100)
```

## Scientific names

Read list of scientific names:

```{r read_taxa_names}
taxa_names <- readr::read_tsv(
  file = here::here(
    "data",
    "interim",
    "timeseries_taxonomic_info.tsv"
  ),
  na = ""
)
```

These names will be added to ranking for better readability.

# Preprocessing

Merge results from GAM and decision rules:

```{r concatenate_gam_decision_rules}
em_list <- append(em_gam_list, em_decision_rules_list)
```


We have now to make a distinction between Belgium and its provinces. Assessment at each of these levels also takes into account the emerging status of the protected areas. In addition, separate assessments can be done for protected areas in specific regions.

We can define a general function to spread columns for ranking. We spread the data frames by model (GAM or decision rules), year and indicator (occurrences or occupancy). We then assign a unique emerging score (if GAM is NA, we choose decision rules). We spread by years and indicators and add scientific names.

```{r spread_em_dfs}
spread_em_dfs <- function(em_list) {
  # Make a data.frame binding all dfs by row
  em_df <-
    dplyr::bind_rows(em_list) %>%
    tidylog::select(-c(dplyr::starts_with("growth"),
                       dplyr::starts_with("dr"))) %>%
    # Spread by model
    tidylog::pivot_wider(
      names_from = model,
      id_cols = c(taxonKey, year, indicator),
      values_from = em_status) %>%
    # Assign unique emerging score (if GAM is NA, choose decision rules)
    tidylog::mutate(em = ifelse(is.na(GAM), decision_rules, GAM),
           method = ifelse(is.na(GAM), "decision_rules", "GAM")) %>%
    # Spread by years
    tidylog::pivot_wider(
      names_from = year,
      values_from = em,
      id_cols = c(taxonKey, indicator),
      names_prefix = paste0("year_")) %>%
    # Spread by indicator
    tidylog::pivot_wider(
      names_from = indicator,
      id_cols = taxonKey,
      values_from = starts_with("year"),
      names_prefix = "em_status_") %>%
    # Add scientific names
    tidylog::left_join(taxa_names, by = "taxonKey")
  em_df
}
```

## Belgium

Select only list elements containing `Belgium` in the name:

```{r em_list_be}
em_list_be <- em_list[grep("Belgium", names(em_list))]
```

Apply the function `spread_em_dfs()` to Belgium:

```{r spread_em_dfs_be}
em_df_be <- spread_em_dfs(em_list_be)
```

To rank taxa equally placed, we use the average of the minimal guaranteed growth of number of occurrences in all Belgium over all the evaluation years.

```{r add_mean_growth_occs_be}
# Add mean growth to em_df
em_df_be <-
  em_df_be %>%
  tidylog::left_join(em_gam_list$gam_occs_Belgium %>%
    tidylog::group_by(taxonKey) %>%
    tidylog::summarize(mean_growth = mean(growth, na.rm = TRUE))) %>%
  tidylog::relocate(mean_growth, .before = canonicalName)
```

A preview (100 rows):

```{r preview_em_df_be}
em_df_be %>%
  tidylog::filter(!is.na(mean_growth)) %>%
  tidylog::slice_head(n = 100)
```

## Regions

Select only list elements containing the region in the name:

```{r em_list_regions}
regions <- list("Flanders", "Wallonia", "Brussels")
names(regions) <- regions
em_list_regions <- purrr::map(regions, 
                              function(r) em_list[grep(r, names(em_list))]
)
```

Apply the function `spread_em_dfs()` to each region:

```{r spread_em_dfs_regions}
em_df_regions <- purrr::map(em_list_regions, spread_em_dfs)
```

To rank taxa equally placed, we use the average of the minimal guaranteed growth of number of occurrences over all the evaluation years.

```{r add_mean_growth_occs_regions}
# Add mean growth to em_df_regions
em_df_regions <-
  em_df_regions %>% purrr::imap(
    function(df, region) {
      occs_column <- paste0("gam_occs_", region)
      df %>%
        tidylog::left_join(em_gam_list[[occs_column]] %>%
          tidylog::group_by(taxonKey) %>%
          tidylog::summarize(mean_growth = mean(growth, na.rm = TRUE))) %>%
        tidylog::relocate(.data$mean_growth, .before = .data$canonicalName)
    }
  )
```

A preview (100 rows) for each region:

```{r preview_em_df_regions}
em_df_regions %>%
  purrr::map(function(df) {
    df %>%
      tidylog::filter(!is.na(mean_growth)) %>% tidylog::slice_head(n = 100)
})
```

## Protected areas within regions

Select only list elements containing natura2000 and the region in the name:

```{r em_list_protected_regions}
protected_regions <- list("natura2000_Flanders", "natura2000_Wallonia", "natura2000_Brussels")
names(protected_regions) <- protected_regions
em_list_protected_regions <- purrr::map(protected_regions, 
                              function(r) em_list[grep(r, names(em_list))]
)
```

Apply the function `spread_em_dfs()` to each regional protected area:

```{r spread_em_dfs_protected_regions}
em_df_protected_regions <- purrr::map(em_list_protected_regions, spread_em_dfs)
```

To rank taxa equally placed, we use the average of the minimal guaranteed growth of number of occurrences over all the evaluation years.

```{r add_mean_growth_occs_protected_regions}
# Add mean growth to em_df
em_df_protected_regions <-
  em_df_protected_regions %>% purrr::imap(
    function(df, region) {
      occs_column <- paste0("gam_occs_", region)
      df %>%
        tidylog::left_join(em_gam_list[[occs_column]] %>%
          tidylog::group_by(taxonKey) %>%
          tidylog::summarize(mean_growth = mean(growth, na.rm = TRUE))) %>%
        tidylog::relocate(.data$mean_growth, .before = .data$canonicalName)
    }
  )
```

A preview (100 rows) for each region:

```{r preview_em_df_protected_regions}
em_df_protected_regions %>%
  purrr::map(function(df) {
    df %>%
      tidylog::filter(!is.na(mean_growth)) %>% tidylog::slice_head(n = 100)
})
```

# Ranking by emerging status

Now we are ready to rank the alien species based on the emerging status.

## Hierarchical ranking

### Belgium

The ranking is based on the highest emerging status. The following priority rules are applied, in order of importance:

1. the more recent, the higher priority is
2. emerging statuses in protected areas are more important than the ones defined over entire Belgium
3. emerging statuses from occupancy are more important than the ones from occurrences
4. the higher average minimal guaranteed growth (#occs/year), the higher priority is

We first define a function to rank the emerging status of species based on the priority rules above. The function takes as input the years, indicators, protection status, regions and the data frame with emerging status scores. It returns a data frame with the ranking of species based on the priority rules. The function is called `hierarchical_ranking()` and is defined as follows:

```{r hierarchical_ranking_function}
hierarchical_ranking <- function(years, indicators, protection_status, regions, df) {
  # Generate the column names dynamically
cols_to_sort <- expand.grid(indicator = indicators, protectionstatus=protection_status, region = regions, year = years) %>%
    mutate(col_name = ifelse(protectionstatus=="entire_region",
                             paste("year", year, "em_status", indicator,region, sep = "_"), 
                             paste("year", year, "em_status", indicator,protectionstatus, region, sep = "_"))) %>%
    pull(col_name)

  # Add mean_growth to the end of the sorting columns
  cols_to_sort <- c(cols_to_sort, "mean_growth")

  # Apply the sorting
  ranking <- df %>%
    arrange(across(all_of(cols_to_sort), desc)) %>%
    tidylog::select(
      taxonKey, canonicalName, kingdom, class,
      all_of(cols_to_sort),
      dplyr::everything()
    )
  ranking
}
```

We apply the function to Belgium and its Natura2000 protected areas:

```{r rank_basic_be}
years <- names(em_df_be) %>%
  grep("^year_\\d+_em_status_", ., value = TRUE) %>%
  gsub("^year_(\\d+)_em_status_.*", "\\1", .) %>%
  as.numeric() %>%
  unique() %>%
  sort(decreasing = TRUE)

indicators <- c("occupancy", "occs")
protection_status<-c("natura2000", "entire_region")
regions <- "Belgium"

ranking_be <- hierarchical_ranking(years, indicators, protection_status, regions, em_df_be)
ranking_be
```


### Regions
The ranking of species at regional levels will be performed in the same way as indicated before. Specifically, species will be sorted according to the following priority rules (in order of importance):

1. the more recent, the higher priority
2. emerging statuses in protected areas are more important than the ones defined over the entire region
3. emerging statuses from occupancy are more important than the ones from occurrences
4. the higher average minimal guaranteed growth (#occs/year), the higher priority is

We apply the function `hierarchical_ranking()` to Flanders, Wallonia and Brussels:

```{r rank_basic_regions}
years <- names(em_df_be) %>%
  grep("^year_\\d+_em_status_", ., value = TRUE) %>%
  gsub("^year_(\\d+)_em_status_.*", "\\1", .) %>%
  as.numeric() %>%
  unique() %>%
  sort(decreasing = TRUE)

indicators <- c("occupancy", "occs")
protection_status<-c("natura2000", "entire_region")

ranking_regions <- purrr::imap(
  em_df_regions, 
  function(em_df, region) hierarchical_ranking(years, indicators, protection_status, region, em_df)
)
ranking_regions
```


### Protected areas within regions
For some analyses, it can be valuable to rank only species that occur in Natura2000 areas at a regional level. The ranking of these species will be performed in a similar way as before, with the exception that they will not be sorted according to point 2: emerging status in protected areas, as these are already the focus of this analysis. Therefore, species will be sorted according to the following priority rules (in order of importance):

1. the more recent, the higher priority
2. emerging statuses from occupancy are more important than the ones from occurrences
3. the higher average minimal guaranteed growth (#occs/year), the higher priority is

We apply the function `hierarchical_ranking()` to Natura2000 areas of Flanders, Wallonia and Brussels: 

```{r rank_basic_protected_regions}
years <- names(em_df_be) %>%
  grep("^year_\\d+_em_status_", ., value = TRUE) %>%
  gsub("^year_(\\d+)_em_status_.*", "\\1", .) %>%
  as.numeric() %>%
  unique() %>%
  sort(decreasing = TRUE)

indicators <- c("occupancy", "occs")
#Given that the focus is already on Natura2000 regions, we have to indicate "entire_region" as protection_status
protection_status<-c("entire_region")

ranking_protected_regions <- purrr::imap(
  em_df_protected_regions, 
  function(em_df, region) hierarchical_ranking(years, indicators, protection_status, region, em_df)
)
ranking_protected_regions
```

## Point strategy

We can also rank using a point strategy reflecting the same ranking strategy as before. To do so we apply gain factors to all emerging status scores using the number of observations in the earliest evaluation year in all Belgium/region as reference (gain factor  = 1).

The following gain factors are applied for Belgium for evaluation years 2020, 2021 and 2022:

| indicator | year | region | gain factor |
| --- | --- | --- | --- | 
occupancy | 2022 | protected areas | 3
observations | 2022 | protected areas | 2.5
occupancy | 2022 | Belgium | 2.5
observations | 2022 | Belgium | 2
occupancy | 2021 | protected areas | 2.5
observations | 2021 | protected areas | 2
occupancy | 2021 | Belgium | 2
observations | 2021 | Belgium | 1.5
occupancy | 2020 | protected areas | 2
observations | 2020 | protected areas | 1.5
occupancy | 2020 | Belgium | 1.5
observations | 2020 | Belgium | 1

The same gain factors are applied for each region:

| indicator | year | region | gain factor |
| --- | --- | --- | --- | 
occupancy | 2022 | protected areas | 3
observations | 2022 | protected areas | 2.5
occupancy | 2022 | entire region | 2.5
observations | 2022 | entire region | 2
occupancy | 2021 | protected areas | 2.5
observations | 2021 | protected areas | 2
occupancy | 2021 | entire region | 2
observations | 2021 | entire region | 1.5
occupancy | 2020 | protected areas | 2
observations | 2020 | protected areas | 1.5
occupancy | 2020 | entire region | 1.5
observations | 2020 | entire region | 1

For protected areas of specific regions, the gain factors are the same, but the ones for protected areas are not applied.

Specifically, the following gain factors are applied:

| indicator | year | region | gain factor |
| --- | --- | --- | --- | 
occupancy | 2022 | protected areas | 3
observations | 2022 | protected areas | 2.5
occupancy | 2021 | protected areas | 2.5
observations | 2021 | protected areas | 2
occupancy | 2020 | protected areas | 2
observations | 2020 | protected areas | 1.5

### Belgium
We first define a function to rank the emerging status of species based on the point strategy explained above. The function takes as input the years, indicators, protection status, regions and the data frame with emerging status scores. It returns a data frame with the ranking of species based on the aforementioned weighting scheme. The function is called `point_ranking()` and is defined as follows:

```{r point_ranking_function}
point_ranking <- function(years, indicators, protection_status, regions, weights, df) {
  # Generate column names dynamically
  cols_to_weigh <- expand.grid(indicator = indicators, protectionstatus=protection_status, region = regions, year = years) %>%
    mutate(col_name = ifelse(protectionstatus=="entire_region",
                             paste("year", year, "em_status", indicator,region, sep = "_"), 
                             paste("year", year, "em_status", indicator,protectionstatus, region, sep = "_"))) %>%
    pull(col_name)
  
  ranking_pts_be <-df %>%
    rowwise() %>%
    mutate(em_pts = sum(c_across(all_of(cols_to_weigh)) * weights, na.rm = TRUE)) %>%
    ungroup() %>%
    dplyr::arrange(
      desc(em_pts),
      desc(mean_growth)) %>%
    tidylog::select(
      taxonKey, canonicalName, kingdom, class,
      em_pts, mean_growth,
      all_of(cols_to_weigh),
      dplyr::everything()
    )
  ranking_pts_be
}
```

We apply the function to Belgium and its Natura2000 protected areas:

```{r rank_basic_be}
years <- names(em_df_be) %>%
  grep("^year_\\d+_em_status_", ., value = TRUE) %>%
  gsub("^year_(\\d+)_em_status_.*", "\\1", .) %>%
  as.numeric() %>%
  unique() %>%
  sort(decreasing = TRUE)

indicators <- c("occupancy", "occs")
protection_status<-c("natura2000", "entire_region")
regions <- "Belgium"
weights<-c(3, 2.5, 2.5, 2, 2.5, 2, 2, 1.5, 2, 1.5, 1.5, 1)

ranking_pts_be <- point_ranking(years, indicators, protection_status, regions,weights, em_df_be)
ranking_pts_be
```
`

### Regions

The same function is applied to each region and its associated Natura2000 protected areas.

```{r rank_by_em_pts_regions}
years <- names(em_df_be) %>%
  grep("^year_\\d+_em_status_", ., value = TRUE) %>%
  gsub("^year_(\\d+)_em_status_.*", "\\1", .) %>%
  as.numeric() %>%
  unique() %>%
  sort(decreasing = TRUE)

indicators <- c("occupancy", "occs")
protection_status<-c("natura2000", "entire_region")
weights<-c(3, 2.5, 2.5, 2, 2.5, 2, 2, 1.5, 2, 1.5, 1.5, 1)

ranking_pts_regions <- purrr::imap(
  em_df_regions, 
  function(em_df_regions, region) point_ranking(years, indicators, protection_status, region, weights, em_df_regions)
)
ranking_pts_regions
```

### Protected areas within regions
To rank species within protected areas of a specific region, we will use the same point strategy. However, as the focus of this analysis is already on protected areas we will apply a slightly different weighting scheme as indicated above.

We apply the function `point_ranking()` to Natura2000 areas of Flanders, Wallonia and Brussels: 

```{r rank_by_em_pts_protected_regions}
years <- names(em_df_be) %>%
  grep("^year_\\d+_em_status_", ., value = TRUE) %>%
  gsub("^year_(\\d+)_em_status_.*", "\\1", .) %>%
  as.numeric() %>%
  unique() %>%
  sort(decreasing = TRUE)

indicators <- c("occupancy", "occs")
#Given that the focus is already on Natura2000 regions, we have to indicate "entire_region" as protection_status
protection_status<-c("entire_region")
weights<-c(3,2.5,2.5,2,2,1.5)

ranking_pts_protected_regions <- purrr::imap(
  em_df_protected_regions, 
  function(em_df_protected_regions, region) point_ranking(years, indicators, protection_status, region, weights, em_df_protected_regions)
)
ranking_pts_protected_regions
```

# Save results

## Final emerging status

We save the final emerging status for Belgium, `em_dfbe`:

```{r save_em_df_be}
readr::write_tsv(em_df_be, 
          file = here::here("data", "output", "emerging_status_Belgium.tsv"),
          na = ""
)
```

the final emerging status for regions, `em_df_regions`:

```{r save_em_df_regions}
purrr::iwalk(
  em_df_regions,
  function(df, region) {
    readr::write_tsv(df, 
              file = here::here("data", "output", paste0("emerging_status_", region, ".tsv")),
              na = ""
    )
  }
)
```

## Rankings

We save both the hierarchical ranking and the ranking based on points strategy for Belgium, its regions, and the protected areas within each region.

### Hierarchical ranking

The ranking for Belgium based on hierarchical strategy is saved in `data/output/ranking_emerging_status_hierarchical_strategy_Belgium.tsv`:

```{r save_ranking_basic_be}
readr::write_tsv(ranking_be,
          file = here::here("data",
                     "output",
                     "ranking_emerging_status_hierarchical_strategy_Belgium.tsv"),
          na = ""
)
```

The ranking for the regions based on hierarchical strategy are saved in `data/output/ranking_emerging_status_hierarchical_strategy_*.tsv` where * is the region:

```{r save_ranking_basic_regions}
purrr::iwalk(
  ranking_regions,
  function(df, region) {
    readr::write_tsv(df,
              file = here::here("data",
                     "output",
                     paste0("ranking_emerging_status_hierarchical_strategy_", region, ".tsv")),
              na = ""
    )
  }
)
```

### Points strategy

Ranking for Belgium based on points strategy is saved in `data/output/ranking_emerging_status_points_strategy_Belgium.tsv`:

```{r save_ranking_pts_be}
readr::write_tsv(ranking_pts_be,
  file = here::here("data",
                    "output", 
                    "ranking_emerging_status_points_strategy_Belgium.tsv"),
  na = ""
)
```

Ranking for the regions based on points strategy are saved in `data/output/ranking_emerging_status_points_strategy_*.tsv` where * is the region:

```{r save_ranking_pts_regions}
purrr::iwalk(
  ranking_pts_regions,
  function(df, region) {
    readr::write_tsv(df,
              file = here::here("data",
                     "output",
                     paste0("ranking_emerging_status_points_strategy_", region, ".tsv")),
              na = ""
    )
  }
)
```
