---
title: "Taxonomic distribution emerging species"
author:
- Damiano Oldoni
- Tim Adriaens
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This document follows the [ranking pipeline](https://trias-project.github.io/indicators/07_occurrence_indicators_modelling.html) to produce summary tables and graphs about the taxonomic distribution of the species defined as emerging (emerging status = 3).

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

Load libraries:

```{r load_libraries}
library(tidyverse) # To do data science
library(tidylog) # To provide feedback on dplyr functions
library(here) # To find files
```

# Get data

Read the tab separated files containing the emerging status output:

```{r read_ranking_emerging_status_hierarchical_strategy}
em_df <- read_tsv(here::here(
  "data",
  "output", 
  "ranking_emerging_status_hierarchical_strategy.tsv"),
  na = ""
)
```

Preview:

```{r preview_em_df}
head(em_df)
```

# Get number of emerging species per class, year, location and variable

First, reshape the data:

```{r select_only_emerging_species}
emerging_df <-
  em_df %>%
  pivot_longer(starts_with("year_"), 
               names_to = "variable",
               values_to = "emerging_status")
```

Preview:

```{r preview_emerging_df}
head(emerging_df)
```

We calculate the number of emerging species (`emerging_status` = 3) 
at class level for each combination of:

- year
- variable: occupancy or number of observations
- area of interest: Belgium or protectred areas

```{r n_emerging}
emerging_df <- 
  emerging_df %>%
  group_by(.data$class, 
           .data$kingdom,
           .data$variable) %>%
  summarise(n_emerging = length(.data$emerging_status[.data$emerging_status == 3]))
```

Add year, variable and area of interest as columns:

```{r add_vars}
emerging_df <- 
  emerging_df %>%
  mutate(
    year = str_extract(.data$variable, pattern = "\\d+"),
    area = str_extract(.data$variable, pattern = "BE$|pa$"),
    variable = str_extract(.data$variable,
                           pattern = "occupancy|observations")) %>%
  mutate(area = recode(.data$area, BE = "Belgium", pa = "Natura2000")) %>%
  relocate(n_emerging , .after = last_col()) %>%
  ungroup() %>%
  arrange(.data$year, desc(.data$n_emerging))
```

Preview:

```{r preview_emergence_df}
head(emerging_df, 20)
```
Save this output file:

```{r save_emerging_df}
write_tsv(
  emerging_df,
  here::here(
  "data",
  "output", 
  "number_of_emerging_taxa_per_class.tsv"),
  na = ""
)
```


# Summaries and graphs

Kingdoms with emerging species:

```{r kingdoms}
kingdoms <- emerging_df %>% distinct(.data$kingdom)
kingdoms
```

Classes with emerging species:

```{r classes}
classes <- emerging_df %>% distinct(.data$class, .data$kingdom)
classes
```
Distribution of classes at kingdom levels:

```{r classes_per_kingdom}
classes %>%
  group_by(.data$kingdom) %>%
  count() %>%
  arrange(desc(.data$n))
```

## Occupancy

### Belgium

```{r emerging_df_BE_occupancy}
emerging_df_BE_occupancy <-
  emerging_df %>%
  filter(.data$area == "Belgium" & .data$variable == "occupancy")
emerging_df_BE_occupancy
```

```{r plot_emerging_df_BE_occupancy}
emerging_plots_BE_occupancy <- map(kingdoms$kingdom, function(x) {
  data <- 
    emerging_df_BE_occupancy %>%
    filter(.data$kingdom == x & .data$n_emerging > 0)
  ggplot(data) +
    geom_col(aes(x = class, y = n_emerging, fill = year), position = "dodge") +
    theme(
    axis.text.x = element_text(
      angle = 315,
      hjust = 0,
      vjust = 1
      )
    )
})
emerging_plots_BE_occupancy
```

Save both table and graph:

```{r save_emerging_BE_occupancy}
write_tsv(
  emerging_df_BE_occupancy,
  here::here(
  "data",
  "output",
  "taxonomic_distribution_emerging_species",
  "n_emerging_taxa_per_class_in_BE_occupancy.tsv"),
  na = ""
)
walk2(
  emerging_plots_BE_occupancy,
  kingdoms$kingdom, 
  function(x, y) {
    ggsave(here::here("data",
                      "output",
                      "taxonomic_distribution_emerging_species",
                      paste0("n_emerging_taxa_per_class_in_BE_occupancy_",
                             y,
                             ".png")),
           x)
})
```

### Natura2000

```{r emerging_df_Natura2000_occupancy}
emerging_df_Natura2000_occupancy <-
  emerging_df %>%
  filter(.data$area == "Natura2000" & .data$variable == "occupancy")
emerging_df_Natura2000_occupancy
```

```{r plot_emerging_df_Natura2000_occupancy}
emerging_plots_Natura2000_occupancy <- map(
  kingdoms$kingdom, function(x) {
    data <- emerging_df_Natura2000_occupancy %>%
      filter(.data$kingdom == x & .data$n_emerging > 0)
    ggplot(data) +
      geom_col(aes(x = class, y = n_emerging, fill = year), position = "dodge") +
      theme(
        axis.text.x = element_text(
          angle = 315,
          hjust = 0,
          vjust = 1
        )
      )
  })
emerging_plots_Natura2000_occupancy
```

Save both table and graph:

```{r save_emerging_Natura2000_occupancy}
write_tsv(
  emerging_df_Natura2000_occupancy,
  here::here(
  "data",
  "output",
  "taxonomic_distribution_emerging_species",
  "n_emerging_taxa_per_class_in_Natura2000_occupancy.tsv"),
  na = ""
)
walk2(
  emerging_plots_Natura2000_occupancy,
  kingdoms$kingdom, 
  function(x, y) {
    ggsave(here::here(
      "data",
      "output",
      "taxonomic_distribution_emerging_species",
      paste0("n_emerging_taxa_per_class_in_Natura2000_occupancy_",
             y,".png")
      ),
      x)
})
```


## Number of observations

### Belgium

```{r emerging_df_BE_obs}
emerging_df_BE_obs <-
  emerging_df %>%
  filter(.data$area == "Belgium" & .data$variable == "observations")
emerging_df_BE_obs
```

```{r emerging_plots_BE_obs}
emerging_plots_BE_obs <- map(kingdoms$kingdom, function(x) {
  data <- emerging_df_BE_obs %>%
    filter(.data$kingdom == x & .data$n_emerging > 0)
  ggplot(data) +
    geom_col(aes(x = class, y = n_emerging, fill = year), position = "dodge") +
    theme(
    axis.text.x = element_text(
      angle = 315,
      hjust = 0,
      vjust = 1
      )
    )
})
emerging_plots_BE_obs
```

Save both table and graph:

```{r save_emerging_BE_obs}
write_tsv(
  emerging_df_BE_obs,
  here::here(
  "data",
  "output",
  "taxonomic_distribution_emerging_species",
  "n_emerging_taxa_per_class_in_BE_observations.tsv"),
  na = ""
)
walk2(
  emerging_plots_BE_obs,
  kingdoms$kingdom, 
  function(x, y) {
    ggsave(here::here(
      "data",
      "output",
      "taxonomic_distribution_emerging_species",
      paste0("n_emerging_taxa_per_class_in_BE_observations_",
             y,".png")
      ),
      x)
})
```


### Natura2000

```{r emerging_df_Natura2000_obs}
emerging_df_Natura2000_obs <-
  emerging_df %>%
  filter(.data$area == "Natura2000" & .data$variable == "observations")
emerging_df_Natura2000_obs
```

```{r plot_emerging_df_Natura2000_obs}
emerging_plots_Natura2000_obs <- map(kingdoms$kingdom, function(x) {
  data <- emerging_df_Natura2000_obs %>%
    filter(.data$kingdom == x & .data$n_emerging > 0)
  ggplot(data) +
    geom_col(aes(x = class, y = n_emerging, fill = year), position = "dodge") +
    theme(
    axis.text.x = element_text(
      angle = 315,
      hjust = 0,
      vjust = 1
      )
    )
})
emerging_plots_Natura2000_obs
```

Save both table and graph:

```{r save_emerging_Natura2000_obs}
write_tsv(
  emerging_df_Natura2000_obs,
  here::here(
  "data",
  "output",
  "taxonomic_distribution_emerging_species",
  "n_emerging_taxa_per_class_in_Natura2000_observations.tsv"),
  na = ""
)
walk2(
  emerging_plots_Natura2000_obs,
  kingdoms$kingdom, 
  function(x, y) {
    ggsave(here::here(
      "data",
      "output",
      "taxonomic_distribution_emerging_species",
      paste0("n_emerging_taxa_per_class_in_Natura2000_observations_",
             y,".png")
      ),
      x)
})
```
