---
title: "Add region information to EEA grids"
author:
- Damiano Oldoni
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This document describes how to add region information to the EEA grid cells of Belgium.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries:

```{r load_libraries}
# geospatial packages
library(sf)
# tidyverse packages
library(dplyr)
library(tidyr)
library(purrr)
# path packages
library(here)
# visualization packages
library(mapview)
```

# Read geospatial data

## EEA grids

We read the EEA grids of Belgium at 1x1, 10x10 and 100x100 km2 resolution:

```{r read_shapefiles_eea_grids}
paths <- list(utm1_bel = here::here("data", "external", "utm1_bel"),
              utm10_bel = here::here("data", "external", "utm10_bel"),
              utm100_bel = here::here("data", "external", "utm100_bel"))
eea_grids <- purrr::map(paths, ~ sf::read_sf(.))
```

Preview for the grid at 10x10 km2 resolution:

```{r preview_grid}
eea_grids$utm10_bel %>% head()
```

## Read Belgian regions and country

We read the spatial data containing information about the Belgian administrative regions:

```{r read_regions_polygons}
exdir <- here::here("data", "external", "Regions_Belgium")
temp <- unzip(zipfile = here::here("data", "external", "Regions_Belgium.zip"), exdir = exdir)
regions_be <- sf::read_sf(exdir)
unlink(exdir, recursive = TRUE)
```

Preview:

```{r preview taxa}
regions_be
```

We merge the regions to create a polygon of Belgium:

```{r create_country}
country_be <- sf::st_union(regions_be) %>%
  st_as_sf() %>%
  mutate(OBJECTID = 4,
         Naam = "Belgium",
         GEWEST = "Belgium") %>%
  rename(geometry = x)
```

Preview:

```{r preview_BE}
country_be
```

Add country information to Belgian region spatial data.frame:

```{r add_BE_to_regions}
regions_be <- bind_rows(regions_be, country_be)
```

Show merged data.frame:

```{r preview_regions_country}
regions_be
```

Transform the Belgian regions and the country to the Coordinate Reference System (CRS) of the grids:

```{r convert_crs}
regions_be <- regions_be %>% 
  sf::st_transform(crs = sf::st_crs(eea_grids$utm1_bel))
```

# Add regional information to cell grids

Do a cell grid belongs to Belgium? DO a cell grid belongs to Flanders and/or Wallonia and/or Brussels? We intersect the cell grids with EEA grids and create four new columns, `isBelgium`, `isFlanders`, `isWallonia` and `isBrussels` with boolean values (`TRUE`, `FALSE`). We create first a function, `add_region_info()`, to allow us to not repeat ourselves (DRY principle) as we need it for the grids at the three resolutions, 1x1, 10x10 and 100x100 km2:

```{r add_cols}
add_region_info <- function(grid) {
  sf::st_intersection(regions_be, grid, sparse = FALSE) %>%
    as_tibble() %>% 
    select(CELLCODE, GEWEST) %>%
    tidyr::pivot_wider(names_from = GEWEST, values_from = GEWEST) %>%
    rename(isBelgium = BELGIUM,
           isFlanders = Vlaams,
           isWallonia = Waals,
           isBrussels = Brussels) %>%
    dplyr::right_join(grid, join_by(CELLCODE)) %>%
    mutate(across(starts_with("is"), ~ if_else(!is.na(.x), TRUE, FALSE))) %>%
    sf::st_as_sf()
}
```

We apply the function to the three grids:

```{r add_regions_to_grids}
eea_grids <- purrr::map(eea_grids, ~ add_region_info(.))
```

Summary for each of the grid:

```{r summary}
purrr::map(eea_grids, function(x) {
  x %>% 
    dplyr::as_tibble() %>%
    dplyr::group_by(isBelgium, isFlanders, isWallonia, isBrussels) %>% 
    dplyr::count()
  }
)
```

Visualize the grid at 10x10 km2 resolution. Click on a grid cell to visualize a.o. the region(s) it belongs to:

```{r visualize}
mapview(eea_grids$utm10_bel)
```

Save the grid as a GeoPackage:

```{r save_as_GeoPackage}
paths <- list(
  utm1_bel = here::here("data", "output", "utm1_bel_with_regions.gpkg"),
  utm10_bel = here::here("data", "output", "utm10_bel_with_regions.gpkg"),
  utm100_bel = here::here("data", "output", "utm100_bel_with_regions.gpkg")
)
purrr::walk2(eea_grids, paths, ~ st_write(.x, .y))
```
